Extras

##EXTRAS


how does age, poverty level, and sex each influence an individual's susceptibility to asthma? It is hypothesized that there will be a significant correlation between asthma and age, asthma and poverty level, but not a very significant correlation between asthma and sex. If each of these factors do influence an individual's susceptibility to asthma, this will result in a p-value of <0.05 after running a t-test.

```{r age and deaths, fig.width = 6, fig.height = 4}
library(ggplot2)

age <- data.frame(
  category = c(" 0-4", " 5-11", "12-17", "18-24", "25-34", "35-64", "65+"),
  value = c(1.4, 2.4, 2.0, 3.8, 6.4, 11.5, 27.1)
)

ggplot(age, aes(x = category, y = value)) +
  geom_bar(stat = "identity", fill = "darkred") +
  labs(
    title = "Asthma Deaths By Age Group",
    x = "Age (Years)",
    y = "Asthma-Related Deaths Per Million",
    caption = "Figure 1. Asthma-related deaths in 2021 occurring by age group per million as reported by the CDC."
  ) +
  scale_y_continuous(
    breaks = seq(0, 30, by = 2),  # show numbers every 2 units 
    minor_breaks = seq(0, 30, by = 1)  # add smaller grid lines between major ones
  )

```

```{r scatter plot, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center'}

#scatter plot of number of individuals with asthma vs asthma prevalence by state

ggplot(asthma_data, aes(x = population, y = prevalence)) +
  geom_point(aes(color=region)) +
  geom_text(aes(label = state_abrv), vjust = -0.5, size = 3)+
  labs(title = "Number of Individuals With Asthma vs. Asthma Prevalence by State",
       x = "Population",
       y = "Asthma Prevalence (%)") +
  theme_minimal()
```