---
title: "Final R Markdown Report: Asthma and Socioeconomic Factors"
author: "Emma Shumway"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

This report _____.

# BACKGROUND

Here is some background: _____.

```{r age and deaths, fig.width = 6, fig.height = 4}
library(ggplot2)

age <- data.frame(category = c(" 0-4", " 5-11", "12-17", "18-24", "25-34", "35-64", "65+"), value = c(1.4, 2.4, 2.0, 3.8, 6.4, 11.5, 27.1))

ggplot(age, aes(x = category, y = value)) +
  geom_bar(stat = "identity", fill = "darkred") + # stat="identity" uses y-values directly
  labs(title = "Asthma Deaths By Age",x = "Age (Years)",y = "Asthma-Related Deaths Per Million")

```

# STUDY QUESTION and HYPOTHESIS

## Question  

How does age, poverty level, and sex each influence an individual’s susceptibility to asthma?

## Hypothesis 

There will be a significant correlation between asthma and age, asthma and poverty level, but not a very significant correlation between asthma and sex.

## Prediction

If house finches are acting as important amplifying hosts, locations with more house finches in the blood host analysis will also be the same locations with higher positive tests for WNV in mosquito pools.

# METHODS

A t-test will test the significance of asthma being present given certain factors.

## Analysis 1: Asthma and Poverty Level
```
```

```
```

## Analysis 2: Asthma and Sex
```
```


```
```
## Analysis 3: Asthma and Age


# DISCUSSION

## Interpretation - Analysis 1

Insert analysis here.

## Interpretation - Analysis 2

Insert analysis here.

## Interpretation - Analysis 3

Insert analysis here.

# CONCLUSION

The data suggests that ____.

# REFERENCES     

1. Name, etc.