---
title: "Warm-Up Mini-Report: Mosquito Blood Hosts in Salt Lake City, UT"
author: "Emma Shumway"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

This report investigates the WNV transmission cycle of birds to mosquitos in Utah, which can be evaluated by analyzing mosquito bloodmeals. House finches were determined to be of interest as a potential amplifying pool for WNV. Bloodmeal samples from Utah were collected, amplified, sequenced, compared to known species' sequences, and evaluated. It was concluded via statistical tests that having more house finches available increases the chances of WNV appearing in mosquito pools, which may aid WNV mosquito abatement efforts in Utah.

# BACKGROUND

The West Nile Virus (WNV) is a potentially serious viral disease that can be found naturally in reservoirs of birds. Mosquitoes are known to transmit the infection to humans and other animals as they take bloodmeals from infected birds (Hartemink et al., 2007). These bloodmeals can be sampled, amplified via PCR, sequenced, and compared to public datasets to map mosquito bloodmeals to host populations. Samples of mosquito bloodmeals taken from across Utah can provide information on transmission cycle patterns that can be used to inform mosquito abatement efforts in the state (White, 2025). Prior research points to the potential significance of house finches in serving as an amplifying source pool for WNV (Kumar et al., 2003), as seen in the bar plot below. The implications of this data serves as a foundation for this report. 


```{r viremia, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# Choose some colors
cols <- c(rainbow(30)[c(10:29,1:5)])  # rainbow colors

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

# STUDY QUESTION and HYPOTHESIS

## Question  

What bird species is acting as WNV amplifying host in Salt Lake City?

## Hypothesis 

Given prior data, there is potential that house finches act as important amplifying hosts of WNV in Salt Lake City. 
  
## Prediction

If house finches are acting as important amplifying hosts, locations with more house finches in the blood host analysis will also be the same locations with higher positive tests for WNV in mosquito pools.

# METHODS

Mosquitoes were collected in traps in the Salt Lake City area. Bloodmeal samples were retrieved, amplified with PCR, analyzed via gel electrophoresis, and sequenced. The National Center for Biotechnology Information's Blast tool was used to compare the resulting sequences to known sequences in a national database to identify bloodmeal sources. RStudio Posit Cloud was then used to visualize and interpret datasets via charts.

## Analysis 1: Bloodmeal Hosts and WNV
```{r first-analysis, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
```

Horizontal plots:
```{r horiz-plot,fig.align='left',fig.width=16,fig.height=7, warning=FALSE}
## import counts_matrix: data.frame with column 'loc_positives' (0/1) and host columns 'host_*'
counts_matrix <- read.csv("./bloodmeal_plusWNV_for_BIOL3070.csv")

## 1) Identify host columns
host_cols <- grep("^host_", names(counts_matrix), value = TRUE)

if (length(host_cols) == 0) {
  stop("No columns matching '^host_' were found in counts_matrix.")
}

## 2) Ensure loc_positives is present and has both levels 0 and 1 where possible
counts_matrix$loc_positives <- factor(counts_matrix$loc_positives, levels = c(0, 1))

## 3) Aggregate host counts by loc_positives
agg <- stats::aggregate(
  counts_matrix[, host_cols, drop = FALSE],
  by = list(loc_positives = counts_matrix$loc_positives),
  FUN = function(x) sum(as.numeric(x), na.rm = TRUE)
)

## make sure both rows exist; if one is missing, add a zero row
need_levels <- setdiff(levels(counts_matrix$loc_positives), as.character(agg$loc_positives))
if (length(need_levels)) {
  zero_row <- as.list(rep(0, length(host_cols)))
  names(zero_row) <- host_cols
  for (lv in need_levels) {
    agg <- rbind(agg, c(lv, zero_row))
  }
  ## restore proper type
  agg$loc_positives <- factor(agg$loc_positives, levels = c("0","1"))
  ## coerce numeric host cols (they may have become character after rbind)
  for (hc in host_cols) agg[[hc]] <- as.numeric(agg[[hc]])
  agg <- agg[order(agg$loc_positives), , drop = FALSE]
}

## 4) Decide species order (overall abundance, descending)
overall <- colSums(agg[, host_cols, drop = FALSE], na.rm = TRUE)
host_order <- names(sort(overall, decreasing = TRUE))
species_labels <- rev(sub("^host_", "", host_order))  # nicer labels

## 5) Build count vectors for each panel in the SAME order
counts0 <- rev(as.numeric(agg[agg$loc_positives == 0, host_order, drop = TRUE]))
counts1 <- rev(as.numeric(agg[agg$loc_positives == 1, host_order, drop = TRUE]))

## 6) Colors: reuse your existing 'cols' if it exists and is long enough; otherwise generate
if (exists("cols") && length(cols) >= length(host_order)) {
  species_colors <- setNames(cols[seq_along(host_order)], species_labels)
} else {
  species_colors <- setNames(rainbow(length(host_order) + 10)[seq_along(host_order)], species_labels)
}

## 7) Shared x-limit for comparability
xmax <- max(c(counts0, counts1), na.rm = TRUE)
xmax <- if (is.finite(xmax)) xmax else 1
xlim_use <- c(0, xmax * 1.08)

## 8) Plot: two horizontal barplots with identical order and colors
op <- par(mfrow = c(1, 2),
          mar = c(4, 12, 3, 2),  # big left margin for species names
          xaxs = "i")           # a bit tighter axis padding

## Panel A: No WNV detected (loc_positives = 0)
barplot(height = counts0,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (-)",
        xlim = xlim_use)

## Panel B: WNV detected (loc_positives = 1)
barplot(height = counts1,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (+)",
        xlim = xlim_use)

par(op)

## Keep the colors mapping for reuse elsewhere
host_species_colors <- species_colors
```

## Analysis 2: WNV Host Significance
```{r second-analysis-or-plot, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
```


```{r}
# second-analysis-or-plot, glm with house finch alone against binary +/_
glm1 <- glm(loc_positives ~ host_House_finch,
            data = counts_matrix,
            family = binomial)
summary(glm1)

#glm with house-finch alone against positivity rate
glm2 <- glm(loc_rate ~ host_House_finch,
            data = counts_matrix)
summary(glm2)
```
# DISCUSSION

## Interpretation - Bloodmeal Hosts and WNV

The plots show the various species of birds found in the mosquito bloodmeals and their counts. While bloodmeals positive and negative for WNV have large numbers of house finch hosts, the WNV positive chart shows over double the number of positive WNV cases compared to the number of negative WNV cases in house finch hosts. This is a good signal that house finches are amplifying pools for WNV. It must be noted though, that this test does not indicate statistical signifance without further analysis.

## Interpretation - WNV Host Significance

These GLM tests showed the significance of the correlation between house finches and positive and negative WNV. The predictor variable is house finch bloodmeals, while the response variable is WNV. The second GLM test showed a slightly positive correlation of ~0.027, and even though the slope isn't very high, it has a highly signifcant p-value (Pr(>/t/)) of 0.0000435. This means that is reasonable to suggest that having more house finches available does increase the chances of WNV appearing in mosquito pools.

# CONCLUSION

The data suggests there is a correlation between house finches and WNV in Utah. This information can be useful for mosquito abatement efforts, where a more targeted focus on house finches may help curb the spread of WNV.

# REFERENCES     

1. Hartemink NA, Davis SA, Reiter P, Hubálek Z, Heesterbeek JA. Importance of bird-to-bird transmission for the establishment of West Nile virus. Vector Borne Zoonotic Dis. 2007 Winter;7(4):575-84. doi: 10.1089/vbz.2006.0613. PMID: 17979541.

2. Komar N, Langevin S, Hinten S, Nemeth N, Edwards E, Hettler D, Davis B, Bowen R, Bunning M. Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerg Infect Dis. 2003 Mar;9(3):311-22. https://doi.org/10.3201/eid0903.020628

3. White, G. BIOL 3070 Lecture. 2025, Oct.

4. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.