---
title: "Final R Markdown Report: Asthma Prevalence and Geographic Population Density"
author: "Emma Shumway"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

Asthma affects over 26 million people in the United States and is influenced by a range of genetic and environmental factors. Understanding geographic patterns in asthma prevalence may help identify contributing risks and inform targeted public health action. This study examined whether asthma prevalence differs across U.S. states and regions and whether these differences are associated with population density. Asthma prevalence and population density data were compiled from recent CDC reports and analyzed using spatial visualizations, summary statistics, ANOVA tests, and a Tukey pairwise comparison. While asthma prevalence varied among states, no consistent geographic pattern was apparent. In contrast, prevalence differed significantly across regions, with the Northeast displaying higher asthma prevalence than the South, North Central, and West in several pairwise comparisons. The Northeast also had the highest average population density; however, neither state nor regional analyses showed a significant relationship between population density and asthma prevalence. These findings suggest that asthma prevalence varies by region but is not explained by population density alone. Additional factors such as demographic or socioeconomic characteristics may better account for regional differences. Further research using multi-year datasets and additional variables is recommended to clarify the underlying drivers of geographic variation in asthma prevalence.

# BACKGROUND

Asthma affects over 26 million people in the U.S. (Most Recent Asthma Data, 2024). The development of asthma is not linked to any single factor; rather, there have been a variety of genetic and environmental factors known to be associated with asthma incidence (Khachi et al., 2014). Effective treatment can only be provided through thorough study of the underlying causal factors, so it is useful to examine available data for trends. It's possible that geographic location contribute's to an individual's susceptibility to asthma. This report attempts to answer the question: do different regions of the U.S. vary in asthma prevalence and mortality rates? It is hypothesized that regions that are more densely populated will experience higher rates of asthma due to pollution from large cities. If this is accurate, we expect statistical testing to show a significant difference in asthma prevalence between regions of the U.S. associated with different levels of population density.


# STUDY QUESTION and HYPOTHESIS

## Question  

Do different states and regions of the U.S. vary in asthma prevalence based on population density?

## Hypothesis 

States/regions that are more densely populated will experience higher rates of asthma due to pollution from large cities.

## Prediction

If this is accurate, it is expected an ANOVA test and/or a Tukey pairwise comparison will show a significant difference (p-value <0.05) in asthma prevalence between states or regions of the U.S., or for the correlation between population density and geographic area.

# METHODS and RESULTS

The methods were designed to compare and visualize the data through multiple means. Recent asthma reports from the CDC were collected for testing (Most Recent Asthma Data, 2024). Asthma prevalence rates between states and between regions were visualized with maps. Spatial hotspots for asthma prevalence were assessed for geographic overlap. Then, asthma prevalence was compared by region with a bar plot. Population denstiy by region was also compared with a box plot. Whether there was significance in this data was tested with ANOVA tests and a Tukey pairwise comparison.

Significant difference was found for asthma prevalence by U.S. regions resulting in a p-value of 0.0192. Additionally, significance was found specifically for the Northeast region compared to each other region individually; South (p-value = 0.00419), North Central (p-value = 0.00558), and West (p-value = 0.02671). A pairwise comparison showed the Northeast region also had the highest average population density (549 per square mile).  Comparisons between the other regions showed no significant differences. No significance was found in the comparison between population density and asthma prevalence.

## Analysis 1: Asthma Prevalence By U.S. State

To evaluate whether asthma prevalence differed between U.S. states, counts for asthma incidence were summarized for each state. Population density was recorded at the same time to input datasets all at once. Asthma prevalence was then visualized by U.S. state using a map and a horizontal barplot to highlight states that potentially have much higher rates of asthma. 

```{r setup and data, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#this code inputs the data for asthma prevalence by state

asthma_data <- data.frame(
  state = c("Alabama","Alaska","Arizona","Arkansas","California","Colorado","Connecticut",
            "Delaware","District of Columbia","Florida","Georgia","Hawaii","Idaho","Illinois",
            "Indiana","Iowa","Kansas","Kentucky","Louisiana","Maine","Maryland","Massachusetts",
            "Michigan","Minnesota","Mississippi","Missouri","Montana","Nebraska","Nevada",
            "New Hampshire","New Jersey","New Mexico","New York","North Carolina","North Dakota",
            "Ohio","Oklahoma","Oregon","Pennsylvania","Rhode Island","South Carolina",
            "South Dakota","Tennessee","Texas","Utah","Vermont","Virginia","Washington",
            "West Virginia","Wisconsin","Wyoming"),
  
  state_abrv = c("AL","AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"),
  
  number = c(378702,58742,555997,243032,2629546,500090,356651,
             78980,60410,1646583,802783,103228,161970,856488,
             574447,238689,237705,377365,351940,146031,499134,633126,
             935666,434473,211055,497216,102617,121412,251193,
             148330,644536,171212,1611027,764909,61944,
             1042354,373879,389831,1033741,117252,368883,
             56630,636322,1760654,266105,68112,671051,667228,
             181395,503318,46973),
  
  prevalence_raw = c("9.6 (0.59)","10.7 (0.58)","9.7 (0.49)","10.5 (0.57)","8.7 (0.39)","10.8 (0.43)","12.4 (0.51)",
                     "9.9 (0.64)","11.1 (0.77)","9.3 (0.53)","9.6 (0.48)","9.1 (0.45)","11.1 (0.51)","8.7 (0.58)",
                     "11.0 (0.39)","9.7 (0.42)","10.7 (0.41)","10.8 (0.68)","10.0 (0.55)","13.1 (0.52)","10.4 (0.38)","11.3 (0.43)",
                     "11.9 (0.44)","9.9 (0.32)","9.4 (0.63)","10.4 (0.49)","11.7 (0.51)","8.1 (0.46)","10.1 (0.79)",
                     "13.1 (0.64)","8.9 (0.43)","10.4 (0.62)","10.3 (0.34)","9.2 (0.62)","10.4 (0.60)",
                     "11.4 (0.38)","12.3 (0.54)","11.5 (0.51)","10.1 (0.67)","13.3 (0.64)","9.0 (0.40)",
                     "8.3 (0.75)","11.7 (0.61)","7.9 (0.37)","11.0 (0.43)","12.9 (0.53)","9.9 (0.46)","10.9 (0.26)",
                     "12.9 (0.61)","10.9 (0.42)","10.5 (0.65)"), 
  
  population = c(5108468, 733406, 7431344, 3067732, 38965193, 5877610, 3617176, 1031890, 678972,
  22610726, 11029227, 1435138, 1964726, 12549689, 6862199, 3207004, 2940546,
  4526154, 4573749, 1395722, 6180253, 7001399, 10037261, 5737915, 2939690,
  6196156, 1132812, 1978379, 3194176, 1402054, 9290842, 2114371, 19571216,
  10835491, 783926, 11785935, 4053824, 4233358, 12961683, 1095962, 5373555,
  919318, 7126489, 30503301, 3417735, 647464, 8715698, 7812880, 1770071,
  5910955, 584057),

 pop_density_sqmile = c(101,1.3,65,59,250,57,747,529,11131,422,192,223,24,226,192,57,36,115,106,45,637,898,178,72,63,90,7.8,26,29,157,1263,17,415,223,11,274,59,44,290,1060,179,12,173,117,42,70,221,118,74,109,6)
)


library(dplyr)
library(stringr)

asthma_data <- asthma_data %>%
  mutate(
    prevalence = as.numeric(str_extract(prevalence_raw, "^[0-9.]+")),
    SE = as.numeric(str_extract(prevalence_raw, "(?<=\\().+?(?=\\))")),
    state_lower = tolower(state),
    region = state.region[match(state, state.name)]
  )

asthma_data <- na.omit(asthma_data)

```

```{r prevalence by state, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=6, fig.height=3}

#Asthma Prevalence By State Map

library(ggplot2)
library(maps)
library(dplyr)

# Map data
states_map <- map_data("state")

# Merge map data with asthma data
map_data <- states_map %>%
  left_join(asthma_data, by = c("region" = "state_lower"))

ggplot(map_data, aes(long, lat, group = group, fill = prevalence)) +
  geom_polygon(color = "gray90", size = 0.3) +  # state borders
  geom_polygon(aes(fill = prevalence), color = "white") +
  coord_fixed(1.3) +
 scale_fill_gradientn(
  name = "Asthma Prevalence (%)",
  colors = c("yellow2", "orange", "red", "red4")
) +
  labs(
    title = "Asthma Prevalence by U.S. State"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

```

```{r plot-state-data, echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=7}
#barplot showing ashtma prevalence per state

# Load required libraries
library(tidyverse)

# Read the CSV file
state_data <- read_csv("state_data.csv")

# Create the plot
ggplot(state_data, aes(x = reorder(State,`Percent With Current Athsma`), y = `Percent With Current Athsma`)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = `Percent With Current Athsma` - SE, ymax = `Percent With Current Athsma` + SE), width = 0.3) +
  coord_flip() +
  labs(
    title = "Asthma Rates by State ",
    x = "State",
    y = "Rate (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 10, margin = margin(r = 5)),
    plot.margin = margin(10, 15, 10, 15)
  )


```
These plots show that while there are differences between states, there is not a clear pattern in this distribution. Additional statistical testing will have to be done to determine significance. 


## Analysis 2: Asthma Prevalence By U.S. Region

To evaluate whether asthma prevalence differed between regions of the U.S., counts for asthma incidence were summarized for each region. Regions were designated based on U.S. census data. Asthma prevalence was then visualized by region using a map and a boxplot to highlight regions that potentially have much higher rates of asthma.

```{r regions by color}
#U.S. regions based on the census

# Get U.S. states map data
states_map <- map_data("state")

# Create a dataframe of states and their regions
state_regions <- data.frame(
  state = tolower(state.name),
  region = state.region
)

# Merge map data with region info
map_data <- states_map %>%
  left_join(state_regions, by = c("region" = "state"))

# Plot
ggplot(map_data, aes(x = long, y = lat, group = group, fill = region.y)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  labs(title = "Four U.S. Census Regions",
       fill = "Region") +
  theme_void()
```

```{r region map, include=FALSE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=6, fig.height=3}
#box plot showing asthma prevalence by region

# U.S. states map data
states_map <- map_data("state")

# Put the states into regions in a dataframe
state_regions <- data.frame(
  state = tolower(state.name),
  region = state.region
)

# Merge map data with region info
map_data <- states_map %>%
  left_join(state_regions, by = c("region" = "state"))

# Plot
ggplot(map_data, aes(x = long, y = lat, group = group, fill = region.y)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  labs(title = "Four U.S. Census Regions",
       fill = "Region") +
  theme_void()
```

```{r box plot state vs asthma prevalence, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=4.75, fig.height=7}

#states and asthma prevalence boxplot

library(ggplot2)

ggplot(asthma_data, aes(x = region, y = prevalence, fill= region)) +
  geom_boxplot() +
  labs(title = "Asthma Prevalence by U.S. Region",
       x = "Region",
       y = "Asthma Prevalence (%)",
       fill = "Region") +
  theme_minimal()

```
The boxplot that follows designated region data shows what appears to be a difference for the northeast region. Additional statistical testing will have to be done to determine significance. 


## Analysis 3: Population Density By U.S. Region

In order to run statistical tests on whether asthma prevalence is correlated with population density within regions, average population density by U.S. region was calculated and then visualized with a boxplot.

```{r}

#average population density by region table

# Define custom region lists
west <- c("washington","oregon","idaho","montana","wyoming","utah",
          "nevada","california","arizona","new mexico","colorado")

north_central <- c("north dakota","south dakota","nebraska","kansas","missouri",
                   "iowa","minnesota","wisconsin","illinois","michigan",
                   "indiana","ohio")

south <- c("texas","oklahoma","louisiana","mississippi","arkansas","tennessee",
           "alabama","georgia","florida","south carolina","north carolina",
           "virginia","kentucky","west virginia","maryland","delaware")

northeast <- c("pennsylvania","new jersey","new york","connecticut",
               "rhode island","massachusetts","vermont","new hampshire","maine")

# Assign custom region
asthma_data <- asthma_data %>%
  mutate(
    region_custom = case_when(
      state_lower %in% west ~ "West",
      state_lower %in% north_central ~ "North Central",
      state_lower %in% south ~ "South",
      state_lower %in% northeast ~ "Northeast",
      TRUE ~ NA_character_
    )
  )

# Calculate average population density by region
avg_density_by_region <- asthma_data %>%
  group_by(region_custom) %>%
  summarise(
    avg_population_density = mean(pop_density_sqmile, na.rm = TRUE)
  )

avg_density_by_region

```

```{r box plot population vs state region, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=4.75, fig.height=7}

#population density by U.S. region

library(ggplot2)

ggplot(asthma_data, aes(x = region, y = pop_density_sqmile, fill= region)) +
  geom_boxplot() +
  labs(title = "Population Density by U.S. Region",
       x = "Region",
       y = "Population Density (square mile)",
       fill = "Region") +
  theme_minimal()

```
Again, the northeast appears to have a significantly higher population density than other region, which is consistent with our hypothesis that regions with higher rates of asthma have higher densities. However, this can only be confirmed through statistical testing.


## Analysis 4: Asthma Prevalence and Population Density

Statistical tests could then be run on the collected data. ANOVA tests were run between the Northwest region (region of interest) and prevalence, then population density and prevalence, then finally population density and prevalence by region. A Tukey pairwise comparison was run between region and prevalence factors to further narrow down comparison between regions of interest.

```{r first analysis ANOVA by region and pairwise comparison}

# ANOVA test for region and prevalence, pairwise comparison

# load necessary libraries

library(emmeans)
library("car")

# create a linear model for US region and asthma prevalence

m2 <- lm(prevalence ~ region, data=asthma_data) 

# perform statistical test 

Anova(m2) 
summary(m2)


# pairwise comparison

pairs(emmeans(m2, "region"))

```

```{r second analysis ANOVA for state population density and asthma prevalence, eval=FALSE}

#ANOVA Test for population density and asthma prevalence by state

library("car")
library("lme4")

# create a linear model for population density per square mile and asthma prevalence by state

m1 <- lm(prevalence ~ pop_density_sqmile, data=asthma_data)

# perform statistical test

Anova(m1)
summary(m1)

```


```{r third analysis ANOVA for region population density and asthma prevalence, eval=FALSE}

# ANOVA test for population density and asthma prevalence by region

library("car")

region_data <- data.frame(
  region = c("North Central", "Northeast", "South", "West" ),
  density = c(107, 549, 204, 60),
  prevalence = c(10.1, 11.7, 10.2, 10.6))

# create a linear model for average population density per region and asthma prevalence

r1 <- lm(prevalence ~ density, data=region_data)

# perform statistical test

Anova(r1)
summary(r1)

```
The ANOVA test for region vs prevalence showed significant p-values (0.00419, 0.00558, 0.02671) for the Northeast region compared to the South, North Central, and West regions respectively. The comparison between the Northeast region and the South and West regions remained significant after a Tukey pairwise comparison, while the comparison with the West region no longer showed significance. The ANOVA tests for population density vs prevalence by state and population density vs prevalence by region did not show significant p-values (0.667, 0.1293, respectively).


# DISCUSSION

Results from this analysis support the hypothesis that rates of asthma vary by region, but do not support the hypothesis that this variation is correlated with population density in those regions.

Significant difference was found for asthma prevalence by U.S. regions resulting in a p-value of 0.0192.  A pairwise comparison showed   The Northeast also had the highest average population density (549 per square mile), which supports the hypothesis that more densely populated regions have a higher asthma prevalence. Comparisons between the other regions showed no significant differences. This leads us to conclude that while there are differences in asthma prevalence by region, it is inconclusive if it depends on population density. Further studies could look at other factors, such as age, sex, or race to explain this difference.


## Interpretation - Analysis 1: Asthma Prevalence By U.S. State

States did not show a clear pattern of difference in asthma prevalence. This does not support the hypothesis that prevalence varies by state, but this must be statistically determined.

## Interpretation - Analysis 2: Asthma Prevalence By U.S. Region

Regions, specifically the northeast, did show differences in asthma prevalence based on visual analysis of the boxplot. The northeast region had a higher median and spread of prevalence. This supports the hypothesis that prevalence varies by region, but again, this must be statistically determined.

## Interpretation - Analysis 3: Population Density By U.S. Region

Regions, once again the northeast, did show differences in population density based on visual analysis of the boxplot. The northeast having a higher median and spread for population density than other regions supports the hypothesis that prevalence varies by region, but this specific correlation must be statistically determined.

## Interpretation - Analysis 4: Asthma Prevalence and Population Density

The ANOVA test for region vs prevalence showed significance for the Northeast region compared to the South, North Central, and West regions respectively. Even after the correction with the Tukey pairwise comparison, the comparison between the Northeast region and the South and West regions remained significant, while the comparison with the West region no longer showed significance. The ANOVA tests for population density vs prevalence by state and population density vs prevalence by region did not show significance. These results support the hypothesis that prevalence varies by region, but cannot conclude that prevalence varies by state or with population density.


# CONCLUSION

The data suggests that asthma prevalence varies by region of the U.S., but not by state. It also suggests that the variation seen by region is not significantly correlated with the population density of the region. However, the northeast region is of interest and warrants further investigation as to the significance seen in its asthma prevalence variance. 

Further investigations could explore additional variables that may be contributing to individual asthma development, such as socioeconomic factors like age, sex, or race. 

There are limitations with this report. Data was collected from one source and within only a one year time frame, which may skew results. It is possible that the arbitary nature of assigning states to regions when there are much geographic variation across regions may reduce accuracy when attempting to compare differences in geography.

# REFERENCES     

1. Asthma and Allergy Foundation of America. (n.d.). Air pollution and asthma. Asthma & Allergy Foundation of America. https://aafa.org/asthma/asthma-triggers-causes/air-pollution-smog-asthma/
2. Most Recent Asthma Data. (2024, November 21). CDC. https://www.cdc.gov/asthma-data/about/most-recent-asthma-data.html
3. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions.
    such as ggplot(), creating the map and troubleshooting error messages. Accessed 2025-11-10.
4. Elias, J. A., Zhu, Z., Chupp, G., & Homer, R. J. (1999, October 15). Airway remodeling in asthma. The Journal of Clinical Investigation. https://www.jci.org/articles/view/8124?..
5. Khachi, H., Meynell, H., & Murphy, A. (2014). Asthma: Pathophysiology, causes and diagnosis. The Pharmaceutical Journal. https://pharmaceutical-journal.com/article/ld/asthma-pathophysiology-causes-and-diagnosis
6. Kudo, M., Ishigatsubo, Y., & Aoki, I. (2013, September 9). Pathology of asthma. Frontiers. https://www.frontiersin.org/journals/microbiology/articles/10.3389/fmicb.2013.00263/full 
7. National Heart, Lung, and Blood Institute. (2025). Asthma. U.S. Department of Health and Human Services. https://www.nhlbi.nih.gov/health/asthma
8. Paciência, I., & Cavaleiro Rufo, J. (2020). Urban-level environmental factors related to pediatric asthm... : Porto Biomedical Journal. https://journals.lww.com/pbj/FullText/2020/02000/Urban_level_environmental_factors_related_to.2.aspx
9. Sockrider, M., & Fussner, L. (2020). What is asthma? | American Journal of Respiratory and Critical Care Medicine. https://www.atsjournals.org/doi/10.1164/rccm.2029P25 
10. Win, P. H., & Hussain, I. (2009, May 22). Asthma triggers: What really matters?. Clinical Asthma. https://pmc.ncbi.nlm.nih.gov/articles/PMC7152189/