---
title: "Final R Markdown Report: Asthma and Socioeconomic Factors"
author: "Emma Shumway"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

This report _____.

# BACKGROUND

Asthma affects over 26 million people in the U.S. (Most Recent Asthma Data, 2024). The development of asthma is not linked to any single factor; rather, there have been a variety of genetic and environmental factors known to be associated with asthma incidence (Khachi et al., 2014). Effective treatment can only be provided through thorough study of the underlying causal factors, so it is useful to examine available data for trends. This report attempts to answer the question: how does age, poverty level, and sex each influence an individual's susceptibility to asthma? It is hypothesized that there will be a significant correlation between asthma and age, asthma and poverty level, but not a very significant correlation between asthma and sex. If each of these factors do influence an individual's susceptibility to asthma, this will result in a p-value of <0.05 after running a t-test.


# STUDY QUESTION and HYPOTHESIS

## Question  

How does age, poverty level, and sex each influence an individual’s susceptibility to asthma?

## Hypothesis 

There will be a significant correlation between asthma and age, asthma and poverty level, but not a very significant correlation between asthma and sex.

## Prediction

If each of these factors do influence an individual's susceptibility to asthma, this will result in a p-value of <0.05 after running a t-test.

# METHODS

A t-test will test the significance of asthma being present given certain factors.

##Data for map
Data for map

```{r setup and data, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

asthma_data <- data.frame(
  state = c("Alabama","Alaska","Arizona","Arkansas","California","Colorado","Connecticut",
            "Delaware","District of Columbia","Florida","Georgia","Hawaii","Idaho","Illinois",
            "Indiana","Iowa","Kansas","Kentucky","Louisiana","Maine","Maryland","Massachusetts",
            "Michigan","Minnesota","Mississippi","Missouri","Montana","Nebraska","Nevada",
            "New Hampshire","New Jersey","New Mexico","New York","North Carolina","North Dakota",
            "Ohio","Oklahoma","Oregon","Pennsylvania","Rhode Island","South Carolina",
            "South Dakota","Tennessee","Texas","Utah","Vermont","Virginia","Washington",
            "West Virginia","Wisconsin","Wyoming"),
  
  state_abrv = c("AL","AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"),
  
  number = c(378702,58742,555997,243032,2629546,500090,356651,
             78980,60410,1646583,802783,103228,161970,856488,
             574447,238689,237705,377365,351940,146031,499134,633126,
             935666,434473,211055,497216,102617,121412,251193,
             148330,644536,171212,1611027,764909,61944,
             1042354,373879,389831,1033741,117252,368883,
             56630,636322,1760654,266105,68112,671051,667228,
             181395,503318,46973),
  
  prevalence_raw = c("9.6 (0.59)","10.7 (0.58)","9.7 (0.49)","10.5 (0.57)","8.7 (0.39)","10.8 (0.43)","12.4 (0.51)",
                     "9.9 (0.64)","11.1 (0.77)","9.3 (0.53)","9.6 (0.48)","9.1 (0.45)","11.1 (0.51)","8.7 (0.58)",
                     "11.0 (0.39)","9.7 (0.42)","10.7 (0.41)","10.8 (0.68)","10.0 (0.55)","13.1 (0.52)","10.4 (0.38)","11.3 (0.43)",
                     "11.9 (0.44)","9.9 (0.32)","9.4 (0.63)","10.4 (0.49)","11.7 (0.51)","8.1 (0.46)","10.1 (0.79)",
                     "13.1 (0.64)","8.9 (0.43)","10.4 (0.62)","10.3 (0.34)","9.2 (0.62)","10.4 (0.60)",
                     "11.4 (0.38)","12.3 (0.54)","11.5 (0.51)","10.1 (0.67)","13.3 (0.64)","9.0 (0.40)",
                     "8.3 (0.75)","11.7 (0.61)","7.9 (0.37)","11.0 (0.43)","12.9 (0.53)","9.9 (0.46)","10.9 (0.26)",
                     "12.9 (0.61)","10.9 (0.42)","10.5 (0.65)"), 
  
  population = c(5108468, 733406, 7431344, 3067732, 38965193, 5877610, 3617176, 1031890, 678972,
  22610726, 11029227, 1435138, 1964726, 12549689, 6862199, 3207004, 2940546,
  4526154, 4573749, 1395722, 6180253, 7001399, 10037261, 5737915, 2939690,
  6196156, 1132812, 1978379, 3194176, 1402054, 9290842, 2114371, 19571216,
  10835491, 783926, 11785935, 4053824, 4233358, 12961683, 1095962, 5373555,
  919318, 7126489, 30503301, 3417735, 647464, 8715698, 7812880, 1770071,
  5910955, 584057),

 pop_density_sqmile = c(101,1.3,65,59,250,57,747,529,11131,422,192,223,24,226,192,57,36,115,106,45,637,898,178,72,63,90,7.8,26,29,157,1263,17,415,223,11,274,59,44,290,1060,179,12,173,117,42,70,221,118,74,109,6)
)


library(dplyr)
library(stringr)

asthma_data <- asthma_data %>%
  mutate(
    prevalence = as.numeric(str_extract(prevalence_raw, "^[0-9.]+")),
    SE = as.numeric(str_extract(prevalence_raw, "(?<=\\().+?(?=\\))")),
    state_lower = tolower(state),
    region = state.region[match(state, state.name)]
  )

asthma_data <- na.omit(asthma_data)

```


##Box plot code

```{r box plot, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=4.75, fig.height=7}

library(ggplot2)

ggplot(asthma_data, aes(x = region, y = prevalence)) +
  geom_boxplot() +
  labs(title = "Asthma Prevalence by U.S. Region",
       x = "Region",
       y = "Asthma Prevalence (%)") +
  theme_minimal()

```


Map
```{r prevalence by state, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=6, fig.height=3}

library(ggplot2)
library(maps)
library(dplyr)

# Map data
states_map <- map_data("state")

# Merge map data with asthma data
map_data <- states_map %>%
  left_join(asthma_data, by = c("region" = "state_lower"))

ggplot(map_data, aes(long, lat, group = group, fill = prevalence)) +
  geom_polygon(color = "gray90", size = 0.3) +  # state borders
  geom_polygon(aes(fill = prevalence), color = "white") +
  coord_fixed(1.3) +
 scale_fill_gradientn(
  name = "Asthma Prevalence (%)",
  colors = c("yellow2", "orange", "red", "red4")
) +
  labs(
    title = "Asthma Prevalence by U.S. State"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

```


## Analysis 1: Asthma and Poverty Level
```{r box plot 2, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=4.75, fig.height=7}

library(ggplot2)

ggplot(asthma_data, aes(x = region, y = prevalence)) +
  geom_boxplot() +
  labs(title = "Asthma Prevalence by U.S. Region",
       x = "Region",
       y = "Asthma Prevalence (%)") +
  theme_minimal()

```

This map shows how the regions were split up

```{r region map, include=FALSE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=6, fig.height=3}

# U.S. states map data
states_map <- map_data("state")

# Put the states into regions in a dataframe
state_regions <- data.frame(
  state = tolower(state.name),
  region = state.region
)

# Merge map data with region info
map_data <- states_map %>%
  left_join(state_regions, by = c("region" = "state"))

# Plot
ggplot(map_data, aes(x = long, y = lat, group = group, fill = region.y)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  labs(title = "Four U.S. Census Regions",
       fill = "Region") +
  theme_void()
```


Scatter Plot Code
```{r scatter plot, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center'}

ggplot(asthma_data, aes(x = population, y = prevalence)) +
  geom_point(aes(color=region)) +
  geom_text(aes(label = state_abrv), vjust = -0.5, size = 3)+
  labs(title = "Number of Individuals With Asthma vs. Asthma Prevalence by State",
       x = "Population",
       y = "Asthma Prevalence (%)") +
  theme_minimal()
```


## Analysis 2: Asthma and Sex
```
```


```
```
## Analysis 3: Asthma and Age
```{r age and deaths, fig.width = 6, fig.height = 4}
library(ggplot2)

age <- data.frame(
  category = c(" 0-4", " 5-11", "12-17", "18-24", "25-34", "35-64", "65+"),
  value = c(1.4, 2.4, 2.0, 3.8, 6.4, 11.5, 27.1)
)

ggplot(age, aes(x = category, y = value)) +
  geom_bar(stat = "identity", fill = "darkred") +
  labs(
    title = "Asthma Deaths By Age Group",
    x = "Age (Years)",
    y = "Asthma-Related Deaths Per Million",
    caption = "Figure 1. Asthma-related deaths in 2021 occurring by age group per million as reported by the CDC."
  ) +
  scale_y_continuous(
    breaks = seq(0, 30, by = 2),  # show numbers every 2 units 
    minor_breaks = seq(0, 30, by = 1)  # add smaller grid lines between major ones
  )

```

Bar Plot showing prevalence per state PART 2 NOAH 


```{r plot-state-data, echo=FALSE, message=FALSE, warning=FALSE, fig.height=10, fig.width=7}
# Load required libraries
library(tidyverse)

# Read the CSV file
state_data <- read_csv("state_data.csv")

# Create the plot
ggplot(state_data, aes(x = reorder(State,`Percent With Current Athsma`), y = `Percent With Current Athsma`)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = `Percent With Current Athsma` - SE, ymax = `Percent With Current Athsma` + SE), width = 0.3) +
  coord_flip() +
  labs(
    title = "Asthma Rates by State ",
    x = "State",
    y = "Rate (%)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 10, margin = margin(r = 5)),
    plot.margin = margin(10, 15, 10, 15)
  )


```
##ANOVA Test
install.packages("tidyverse")
install.packages("readxl")
install.packages("car")
install.packages("lme4")
install.packages("emmeans")
install.packages("ggplot2")

```{r anova-test, message=FALSE, warning=FALSE}
# --- Load required packages (must already be installed) ---
library(tidyverse)   # data wrangling and ggplot2
library(readxl)      # reading Excel files
library(car)         # ANOVA tests
library(lme4)        # linear and mixed models
library(emmeans)     # estimated marginal means
library(ggplot2)     # graphics

# 1. First, we create a simple linear model. We name the model m1.
m1 <- lm(prevalence ~ pop_density_sqmile, data=asthma_data) # here, my response (dependent) variable is prevalence, and the predictor (independent) variable is population density. In words, this model is asking, "does the data in column pop_density_sqmile explain variance in column prevalence, within the dataset called asthma_data?".

# 2. Next, we conduct our statistical test: an analysis of variance.
Anova(m1) # This tells us whether there is a significant effect of population density on asthma prevalence, and provides us the ingredients for our statistical reporting. Cool!
summary(m1) # This gives other important ingredients to report from our statistics, such as the amount of variance explained by our independent variable; the degrees of freedom; and our test statistic. The important information to report is at the bottom of this output.

```


# DISCUSSION

## Interpretation - Analysis 1

Insert analysis here.

## Interpretation - Analysis 2

Insert analysis here.

## Interpretation - Analysis 3

Insert analysis here.

# CONCLUSION

The data suggests that ____.

# REFERENCES     

1. Khachi, H., Meynell, H., & Murphy, A. (2014). Asthma: Pathophysiology, causes and diagnosis. The Pharmaceutical Journal. https://pharmaceutical-journal.com/article/ld/asthma-pathophysiology-causes-and-diagnosis

2. Most Recent Asthma Data. (2024, November 21). CDC. https://www.cdc.gov/asthma-data/about/most-recent-asthma-data.html

